---
title: "Buffon's needle"
author: "Paul M"
date: "2/8/2021"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Buffon's needle - solution

This script allows you to estimate pi using the logic of Buffon's needle.

As ever, remember to set the random number seed so that you can reproduce your output if you end up needing to hunt bugs. We also define some other control variables.
```{r}
set.seed(2053)   
LineDistance <- 1  # The distance between the lines
NumberOfTrials <- 50000    # how many needles to throw
```


We will compare performance for needles of different lengths, so now we define the range of lengths and set up a variable in which to store our pi estimates. It's generally best to 
predeclare your memory if you can. Code can run slowly if you keep declaring new memory (e.g. using cbind or rbind)
```{r}
StartLength <- 0.1 # shortest needle considered
EndLength <- 2 # longest needle considered
StepSize <- 0.1 # the differences between the different lengths of needle we will consider
OurLengths <- seq(from=StartLength,to=EndLength,by=StepSize)
NumberOfLengths <- length(OurLengths) # How many needle lengths will we consider
MyPi <- rep(0,NumberOfLengths)  # A place to store our esitmate of pi
```

Now we perform the Buffon's needle experiment for our range of needle lengths (one at a time)

```{r}
IterationCounter <- 0 # a variable we will use to keep track of where we are at for storing answers
for (NeedleLength in OurLengths){
  IterationCounter <- IterationCounter+1
  NumberOfNeedlesThatCrossALine<-0   # this will count how often the needle crosses a line
  # do the experiment - here's one way to do it
  for (i in 1:NumberOfTrials){
    # generate the angle at which the needle falls - R measure angles in radians, so for the code we need to use pi (a pre-defined variable in R)
    Angle <- runif(1,0,pi)
    Height=NeedleLength*sin(Angle)

    # it will cross the a line with prob Height/LineDistance
    PCrossLine <- Height / LineDistance
    p <- runif(1)
    if (p<PCrossLine){ NumberOfNeedlesThatCrossALine <- NumberOfNeedlesThatCrossALine + 1 }
  }

  ProportionOfPointsThatCrossLine <- NumberOfNeedlesThatCrossALine/NumberOfTrials
  MyPi[IterationCounter]<-NeedleLength*2/(ProportionOfPointsThatCrossLine*LineDistance)
  cat("\nNeedle length ",NeedleLength,": estimate of pi is ",MyPi[IterationCounter])
  # and then do the math to produce you estimate of pi - yours to write

}
```


Let's graph our estimates to see how they look
```{r}
plot(x=seq(from=StartLength,to=EndLength,by=StepSize),y=MyPi,main="Estimates of pi when lines are distance 1 apart",xlab="needle length",ylab="Estimated value of pi",pch="*")
abline(pi,0,col="red",lty=2)
```


They look good for shorter needle lengths, but something goes wrong when the needle length is greater than 1. What is the problem there?

If you are feeling fancy, you could write some code to plot your needles and the table as you perform the iterations.



