---
title: "ConditionedExpos"
author: "Paul M"
date: "12/4/2020"
output: html_document
---

Our task here is to produce exponential random variables, using the Universality of the Uniform, and conditioned on them being >y, for some y. We wil use a rejection method to make sure they satisfy the latter condition.

Set-up your global variables...
```{r}
set.seed(999)  # remember to set your random number seed

# set up the variables we need for this task
NumberOfRVs <- 10000  # How many rvs we want to generate
y <- 1    # the value y such that we condition on x>y
lambda <- 1   # the exponential parameter
OurExpos <- rep(0,NumberOfRVs)  # a place to put the answers
```


Now generate Uniform rvs and transform them to become exponentials. Then check whether they are >y.
(So write a while loop to check that condition, for example)
```{r}
# Set up a loop to generate the rvs
for (i in 1:NumberOfRVs){
  
  # generate a unirm
  u<-runif(1,0,1)

  # Transform u to get an expo rv
  ConditionedExpRV<- (-1/lambda)*log(u)
  # It is >y?
  while (ConditionedExpRV < y){
    u<-runif(1,0,1)
    ConditionedExpRV<- (-1/lambda)*log(u)
  }
  # When you leave this loop you should have an expo(lambda) rv that is >y
  
  #Store the value of ConditionedExpRV
  OurExpos[i] <- ConditionedExpRV
}
```

Now construct the plots you were asked for
```{r}
hist(OurExpos,breaks=25,main=c("Exponentials conditioned on being bigger than ",y),col="red",prob=TRUE)
curve(dexp(x-y, rate=lambda), add=TRUE,lwd=2)
```
