---
title: "MetropolisForNormalsWorking"
author: "Paul M"
date: "1/29/2021"
output:
  pdf_document: default
  html_document: default
---


Here's the code for the Metropolis algorithm:

We assume we pass it the following parameters:

* "Mean": the mean of the normal distribution from which we wish to sample

* "SD": the std. dev. of the normal distribution from which we wish to sample

* "EndPt": We will start the algorithm at a point chosen uniformly at random in the range  [-EndPt, +EndPt] 

* "NumberOfIts": We will run the Metrop algorithm for NumberOfIts iterations

```{r Metrop}
MHNormal <- function(Mean,SD,EndPt,NumberOfIts){ # Mean and SD are the Mean and Std. dev. of the Normal,
	#We simulate over the range [-EndPt, +EndPt] and run NumberOfIts iterations.
	NormalXs <- rep(-9,NumberOfIts)  # declare a vector to store the x-values we generate during the course of the algorithm
	# you need to start somewhere: sample a random number, x, between +EndPt and -EndPt
	x = runif(1, -EndPt, EndPt)

		for (i in 1:NumberOfIts){    # each iteration performs one step of the MCMC process
		# propose a new x-value (xprime) by adding a Unif(-0.5,0.5) (say) random variable to the current value
	  if (runif(1)<0.9){
	    xprime  <-  x + runif(1,0,1)
	  }else{
	    xprime  <-  x - runif(1,0,1)
	  }
		# decide whether to move to the new value - 
			# Calculate Metropolis term (the ratio of the likelihoods of the two points, i.e., H <- f(x')/f(x))
			H <- min(1,dnorm(xprime,Mean,SD)/dnorm(x,Mean,SD))
			p <- runif(1,0,1) # generate a random number and then decide whether or not to move to the new value
		if (p<H){ 
			x <- xprime      # move to x'
		} # we dont need an else, since we just keep the existing x in that case
			
			# update the vector that you keep each iterations x-value inNormalXs
			NormalXs[i] <- x
		
		# every now and again, plot a histogram of the accepted values
    HistBars  <-  seq(-EndPt, EndPt, 0.1)
		}
	 hist(NormalXs,breaks=HistBars) 
	return (NormalXs)  # returns the vector of x-values
}
```

Let's test it:
```{r test}
set.seed(572)
MyMH <- MHNormal(0,1,10,20000)
```

Let's look at what we got:
```{r LookSee}
plot(MyMH,type='l')
plot(MyMH[1:1000],type='l')
acf(MyMH,lag.max=200)
```

Is it normal? Test using a QQ plot
```{r qq}
qqnorm(MyMH,pch='.')
abline(0,1,col="red",lty=2)
qqnorm(MyMH[-(1:1000)],pch='.')
abline(0,1,col="red",lty=2)
```


Note that if we remove the first 1000 iterations it starts to look more normal. This is a common feature of metropolis algorithms (and of MCMC). Initially, the points are influenced by where you happened to start from. So we typically remove a "burn-in" period before we start to sample. Formally, this is so that we are only sampling values once the Markov chain reaches "stationarity".
