---
title: "NR"
author: "Paul M"
date: "12/7/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

First, we define some functions to work with
```{r }
F1<-function(x){
		return(c(x^2,2*x))
}

F2<-function(x){
		return(c(sin(x),cos(x)))
}

F3<-function(x){
	y<-(x-2)^3-6*x
	deriv<-3*(x-2)^2-6
	return(c(y,deriv))
}

F4<-function(x){
	return(c(cos(x)-x,-1*sin(x)-1))
}
```

Now we write a function to perform the Newton-Raphson method...
```{r}
NewtonRaphson <- function(func,StartingValue,Tolerance,MaxNumberOfIterations){
	Deviation <- 1000   # an arbitrary big number
	i <- 0
	X <- StartingValue
	Y <- func(X)[1]    
	segments(X,0,X,Y,lty=2,lwd=2,col="blue")
	text(X-0.2,-1.2,"X1",cex=0.75)
	
	#Set up a while loop until we hit the required target accuracy or the max. number of steps
	while ((i<MaxNumberOfIterations) && (Deviation>Tolerance))
	{
		# Find the next X-value using Newton-Raphson's formula
		Z <- func(X)
		NewX <- X-Z[1]/Z[2]

		if ((Z[1]=="NaN") || (Z[2]=="NaN")){
			cat("Function or derivative not defined error.\n")
			break
		}

		# annotate
		segments(X,Y,NewX,0,lty=2,lwd=2,col="blue")
		segments(NewX,0,NewX,func(NewX)[1],lty=2,lwd=2,col="blue")
		#browser()
		text(NewX-0.2,-1.2,paste("X",i+2,sep=""),cex=0.75)

		# calculate accuracy: |f(x)-0|
		X <- NewX
		Y <- func(X)[1]    
		Deviation <- abs(Y)

		i <- i+1
		cat(paste("\nIteration ",i,":   X=",X,"  Y=",Y))
	}

	# output the result
	if (Deviation<Tolerance){
		cat(paste("\nFound the root point: ",X, "after ", i, "iterations"))
		}else{
		cat(paste("\nConvergence failure. Deviation: ",Deviation, "after ", i, 	"iterations"))}
}
```


Now let's run the required analyses.

First $f(x)=x^2$
```{r}
curve(x^2,-3,10)
NewtonRaphson(F1,8,1e-3,40)
```

Then $f(x)=(x-2)^3-6*x$:
```{r}
curve((x-2)^3-6*x,-5,10)
NewtonRaphson(F3,8,1e-3,40)
```

Are there any starting values for which it struggles with that function?

Next we try $f(x)=sin(x)$:
```{r}
curve(sin(x),-5,20)
NewtonRaphson(F2,8,1e-3,40)

curve(sin(x),-5,200)
NewtonRaphson(F2,7.86,1e-3,40)
```

The issue there is that the root you find depends upon the point you start at (in a fairly unpredictable way). We will exploit this fact later to draw some art.
Note also that the second plot of sin(x) looks rather weird. What is going on there?


Now for $f(x)=cos(x)-x$:
```{r}
curve(cos(x)-x,-20,20)
NewtonRaphson(F4,8,1e-3,40)
abline(h=0)

curve(cos(x)-x,-200,200)
NewtonRaphson(F4,8,1e-3,40)
abline(h=0)
```

From a distance this function looks well-behaved (2nd plot), but if we zoom in we see that it is a bit like a set of stairs, and your path to the root can be rather strange is you happen to start at a point at which the tangent is nearly horizontal.
