---
title: "RegressionFitter"
author: "Paul M"
date: "2/25/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Solution to Regression Problem

This algorithm is a partial solution to fitting a regression using optimization.
for linear regression we could iterate the following steps:

1. Test whether the sum of squared residuals [SS] descreases as you increase(decrease)the slope
1a. Increase (or decrease) the slope in steps of size epsilon until the SS stops decreasing.
2. Now test whether the SS decreases as we increase or decrease the intercept.
2a. Increase (or decrease) the slope in steps of size 100*epsilon until the SS stops decreasing.
3. Go back to 1.

So this is like a steepest ascent algorithm in which we force consecutive directions to be perpendicular.

The same idea would works, in principle, for cubic regression, but now we cycle through changing the intercept, slope, squared term, and cubic term.

This partial solution completes just one such cycle, but seems to work pretty well. 

First let's get set up...
```{r}
#filestr<-sprintf("RegressionInput.txt")
filestr<-sprintf("CubicRegressionInput.txt")
Coordinates<-read.table(filestr,header=F)
Residuals<-Coordinates[,1]
Reg<-lm(Coordinates[,2]~Coordinates[,1])

DegreeOfRegression<-3  # what sort of regression do we want? 1=linear; 2=quadratic; 3=cubic

outfile1<-file("RegressionSteps.txt")

#How many points do we have
NumberOfPoints<-length(Coordinates[1,])
```

And now run the algorithm. It takes a while because of all the plotting.
```{r}

plot(Coordinates,pch=46)

# Pick an arbitrary place to start - we might also choose a random slope and intercept
#initialize intercept
intercept<- -2000
#initialize slope
slope<-0
quadraticterm<-0
cubicterm<-0

# jiggle factors - these controls how much we change the slope by at each step
# the first element is for the intercept, the second is for the linear term, the third is for the quadratic term, etc.
epsilon<-rep(0,4)
epsilon[1]<-1
epsilon[2]<-0.1
epsilon[3]<-0.02
epsilon[4]<-0.01

# this function will calculate the sum of squared residuals
CalculateSoS<-function(Coords,slp,inter,quadratic,cubic){
	SS<-0
	Residuals<-Coords[,2]-(inter+slp*Coords[,1]+quadratic*Coords[,1]^2+cubic*Coords[,1]^3)
	SqResiduals<-Residuals*Residuals
	SS<-sum(SqResiduals)
	return (SS)
}

# keep track of by how much we have reduced the Sum of squared residuals
SSred<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)

# record the starting configuration
Recorder<-cbind(intercept,slope,quadraticterm,cubicterm,SSred)
cat(paste("starting configuration",intercept,slope,SSred))

# Do least sum of squares fit using steepest-ascent-like algorithm
count<-0

while (SSred>0.1){
    count<-count+1
	# what is the current SS resids?
	currentSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)

	if (DegreeOfRegression>2){
		# test the coefficient of the cubic term
		cubicterm<-cubicterm+epsilon[4]
		newSS1<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		cubicterm<-cubicterm-2*epsilon[4]
		newSS2<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		cubicGradient<-(newSS1-newSS2)/(2*epsilon[4])
		cubicterm<-cubicterm+epsilon[4]

		# now move in this direction until the SS stops decreasing
		SSdecrease <- 1
		#cat(paste("\nnew direction",count))
		#Move in the right direction
		while ((SSdecrease>0)&&(count<100)){
			OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
			if (newSS1>newSS2)
				cubicterm<-cubicterm-epsilon[4]
			else
				cubicterm<-cubicterm+epsilon[4]
			NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
			SSdecrease<-OldSS-NewSS
			cat(paste("\nchanging cubic: ",intercept,slope,quadraticterm,cubicterm,OldSS,NewSS,SSred))#
			Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))
			# pause for a bit
			Start.Time<-Sys.time()
			while (Sys.time()<Start.Time+0.1){}
			plot(Coordinates,pch=46,main="My Regression")
			abline(Reg)
			curve(intercept+slope*x+quadraticterm*x^2+cubicterm*x^3,col="red",add=TRUE)
		}
		OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		if (newSS1>newSS2)
			cubicterm<-cubicterm+epsilon[4]
		else
			cubicterm<-cubicterm-epsilon[4]
		NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))
	}


	if (DegreeOfRegression>1){
		# test the coefficient of the quadratic term
		quadraticterm<-quadraticterm+epsilon[3]
		newSS1<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		quadraticterm<-quadraticterm-2*epsilon[3]
		newSS2<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		QuadraticGradient<-(newSS1-newSS2)/(2*epsilon[3])
		quadraticterm<-quadraticterm+epsilon[3]

		# now move in this direction until the SS stops decreasing
		SSdecrease <- 1
		#cat(paste("\nnew direction",count))
		#Move in the right direction
		while ((SSdecrease>0)&&(count<100)){
			OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
			if (newSS1>newSS2)
				quadraticterm<-quadraticterm-epsilon[3]
			else
				quadraticterm<-quadraticterm+epsilon[3]
			NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
			SSdecrease<-OldSS-NewSS
			cat(paste("\nchanging quadratic: ",intercept,slope,quadraticterm,cubicterm,OldSS,NewSS,SSred))#
			Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))
		}
		OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		if (newSS1>newSS2)
			quadraticterm<-quadraticterm+epsilon[3]
		else
			quadraticterm<-quadraticterm-epsilon[3]
		NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))
	}


	# test the slope
	slope<-slope+epsilon[2]
	newSS1<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
	slope<-slope-2*epsilon[2]
	newSS2<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
	SlopeGradient<-(newSS1-newSS2)/(2*epsilon[2])
	slope<-slope+epsilon[2]

	# now move in this direction until the SS stops decreasing
	SSdecrease <- 1
	#cat(paste("\nnew direction",count))
	#Move in the right direction
	while ((SSdecrease>0)&&(count<100)){
		OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		if (newSS1>newSS2)
			slope<-slope-epsilon[2]
		else
			slope<-slope+epsilon[2]
		NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		SSdecrease<-OldSS-NewSS
		cat(paste("\nchanging slope: ",intercept,slope,quadraticterm,cubicterm,OldSS,NewSS,SSred))#
		Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))
	}
	OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
	if (newSS1>newSS2)
		slope<-slope+epsilon[2]
	else
		slope<-slope-epsilon[2]
	NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
	#cat(paste("\nchanging slope. New slope: ",intercept,slope,OldSS,NewSS))#
	Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))


	# test the intercept
	intercept<-intercept+epsilon[1]
	newSS1<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
	intercept<-intercept-2*epsilon[1]
	newSS2<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
	InterceptGradient<-(newSS1-newSS2)/(2*epsilon[1])
	intercept<-intercept+epsilon[1]
	SSdecrease<-1
	#Move in the right direction
	while ((SSdecrease>0)&&(count<100)){
		OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		if (newSS1>newSS2)
			intercept<-intercept-epsilon[1]
		else
			intercept<-intercept+epsilon[1]
		NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
		SSdecrease<-OldSS-NewSS
		cat(paste("\nchanging intercept: ",intercept,slope,quadraticterm,cubicterm,OldSS,NewSS,SSred))#
		Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))
	}
	OldSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)
	if (newSS1>newSS2)
		intercept<-intercept+epsilon[1]
	else
		intercept<-intercept-epsilon[1]
	NewSS<-CalculateSoS(Coordinates,slope,intercept,quadraticterm,cubicterm)

	#cat(paste("\nChanging intercept. New int: ",intercept,slope,OldSS,NewSS))#
	Recorder<-rbind(Recorder,cbind(intercept,slope,quadraticterm,cubicterm,NewSS))
	SSred<-currentSS-NewSS
	# pause for a bit
	Start.Time<-Sys.time()
	while (Sys.time()<Start.Time+0){}
	plot(Coordinates,pch=46,main="My Regression")
	abline(Reg)
	curve(intercept+slope*x+quadraticterm*x^2+cubicterm*x^3,col="red",add=TRUE)


}
write(Recorder[,],outfile1,ncolumns=DegreeOfRegression+2) ## output our path
close(outfile1)
```

Now store the final plot to file...
```{r}
str<-sprintf("RegressionFitter1_Output.pdf")
pdf(str)
op <- par(mfrow = c(1,1)) # 2 x 2 pictures on one plot
plot(Coordinates,pch=46,main="My Regression")
abline(Reg)
curve(intercept+slope*x+quadraticterm*x^2+cubicterm*x^3,col="red",add=TRUE)
#plot(Coordinates,pch=46,main="My Regression",abline(intercept,slope))
#abline(intercept,slope)
dev.off()
```
