---
title: "Regression Via SA"
author: "Paul M"
date: "3/2/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This algorithm fits a regression using optimization. I am going to show an example that uses a 
built-in R simulated annealing algorithm in the R package "stats"
```{r libs}
library(stats)
```

We'll start by trying it on linear regression and then see how far we can push it.
start by reading the data and using R's bult-in regression function to see what the 
answer should be:
```{r}
filestr<-sprintf("RegressionInput.txt")
Coordinates<-read.table(filestr,header=F)
Residuals<-Coordinates[,1]
Reg<-lm(Coordinates[,2]~Coordinates[,1])
plot(Coordinates,pch=46,main="Data with regression line from lm()")
abline(Reg$coefficients[1],Reg$coefficients[2],col="blue",lty=2)
print(Reg)
```

So we should get an intercept of  53.915 and a slope of 1.952

First we need to define the functions that will calculate the sum of the residuals for any given fitted line or cubic curve. 

The arguments are as follows:
* Coords: the coordinates of the points we are trying to put a line through
* inter: the intercept
* slp: the slope
* quadratic: the quadratic coefficient
* cubic: the cubic coefft.
(We will be fixing the last two to 0 for the linear regression.)

```{r SoS}
CalculateLinearSoS <- function(params){
  inter <- params[1]
  slp <- params[2]
	SS <- 0
	Residuals <- Coordinates[,2]-(inter+slp*Coordinates[,1])
	SqResiduals <- Residuals*Residuals
	SS <- sum(SqResiduals)
	return (SS)
}

CalculateCubicSoS <- function(params){
  inter <- params[1]
  slp <- params[2]
  quadratic <- params[3]
  cubic <- params[4]
	SS <- 0
	Residuals <- Coordinates[,2]-(inter+slp*Coordinates[,1]+quadratic*Coordinates[,1]^2+cubic*Coordinates[,1]^3)
	SqResiduals <- Residuals*Residuals
	SS <- sum(SqResiduals)
	return (SS)
}
```

Now let's try fitting a line using simulated annealing using the function "optim", from the "stats" library, which has a simulated annealing option (along with many others)
```{r SA}
SAresult <- stats::optim(fn = CalculateLinearSoS, par = c(0,0), method = "SANN", control = list(tmax = 500, reltol = 0.1, temp = 50, trace = TRUE))
print(SAresult)
plot(Coordinates,pch=46,main="SA Regression result (red) and lm() fitted line (blue)")
abline(Reg,col="blue",lty=2)
curve(SAresult$par[1]+SAresult$par[2]*x,col="red",add=TRUE)
```
We see that we get almost identical results.


Now let's try the cubic data.
```{r}
filestr<-sprintf("CubicRegressionInput.txt")
Coordinates<-read.table(filestr,header=F)
Residuals<-Coordinates[,1]
SAresult <- stats::optim(fn = CalculateCubicSoS, par = c(0,0,0,0), method = "SANN", control = list(tmax = 500, reltol = 0.1, temp = 50, trace = TRUE))
print(Reg)
plot(Coordinates,pch=46,main="SA Cubic Regression result")
curve(SAresult$par[1]+SAresult$par[2]*x+SAresult$par[3]*x^2+SAresult$par[4]*x^3,col="red",add=TRUE)

```

Again, we appear to get a good fit.

Further work: 
* Can you write a single sum of squares function that could handle fitting a polynomial of degrees both 1 and 3?).
* How is performance affected if you change some of the parameters for the simulated annealing algorithm?
