---
title: "UrnVersion2_solution"
author: "Paul M"
date: "2/17/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Solution to Urn version 2, counting how many non-black colors there are at the end

This is an example answer for lab task 4 of week 6. Knit it to an html to see
the answers. (I get a Latex error when I knit this to pdf, for reasons I do not see. Also, you will get an error if you just try to "run" this code. It has to be "knit", for reasons I do not yet understand.)

So the goal is to simulate an Urn that contains a black ball and look at
how may different colors there are in it at the end.

This particular code is written in "object-oriented" style. 
This is something that R is not really designed for, but many of us who learned
languages like C++ are used to writing object-oriented code, so I wanted to see
how well R could mimic that style of coding.

As ever, we start with a few global variables...
```{r globals}
set.seed(87)
ColorBreaks <- seq(0.5,10.5,1)
```

We also use a library that allows R to display a "progress bar" that shows
how much longer the code is likely to run for.
```{r progress}
#install.packages("progress")
library(progress)
```


Now, the object oriented bit! We set up a giant function to represent the things in the Urn and 
all the operations we will need to perform on them.
```{r urn}
Urn <- function(StartingConfig,FinalNumberOfBalls,UseMutationBall,WeightOfMutationBall){
  UrnThings <- list(
    # set up the elements of the Urn
    Balls = mat.or.vec(1,FinalNumberOfBalls),    # Note that we need to use "=" here rather than "<-"
    InitialNumberOfBalls = length(StartingConfig),
    UseMutationBall = UseMutationBall,
    MutationBallWeight = WeightOfMutationBall,
    StartingConfiguration = StartingConfig
  )
  
  # A sanity check
  if (FinalNumberOfBalls<length(StartingConfig)){
    cat("Too many balls in starting configuration. Exit.")
  }
  
  # Initialize the balls in the Urn to start with
  Nballs <- length(StartingConfig)
  for (i in 1:Nballs){
    UrnThings$Balls[i] <- StartingConfig[i]
  }
  
  ###############################################################################
  ### Class functions	- these are the functions that will operate on the Urn  ###
  ###############################################################################
  
  UrnThings$DrawFromUrn<-function(NumBalls){
    r<-sample(1:NumBalls,1)
    return (UrnThings$Balls[r])
  }
  
  UrnThings$RunUrnModel<-function(){
    HowManyBallsDoWeNeed<-length(UrnThings$Balls)
    Nballs<-length(UrnThings$StartingConfiguration)
    # set the starting configuration
    for (i in 1:length(StartingConfig)){
      UrnThings$Balls[i]<-StartingConfig[i]
    }
    # how many colors do we have already?
    iColorCounter <- length(apply(UrnThings$Balls,1,unique)) # this is going to keep track of how many new colors we use
    while (Nballs < HowManyBallsDoWeNeed){
      # do we pick the black ball (if there is one)?
      if (UrnThings$UseMutationBall==1){   # this is not actually represented by a ball in the urn
        p<-runif(1,0,1)
        if (p<UrnThings$MutationBallWeight/(UrnThings$MutationBallWeight+Nballs)){
          #  we picked the black ball
          WhichToCopy<-0    # O codes the black ball
        }else{
          WhichToCopy<-UrnThings$DrawFromUrn(Nballs)
        }
      }else{
        WhichToCopy<-UrnThings$DrawFromUrn(Nballs)
      }
      if (WhichToCopy==0){    # 0 is used to code the black ball
        # change the color of some other ball
        #IndexOfBall<-ceiling(runif(1,0,1))   # PM changed this on 2/17/21
        IndexOfBall <- sample(1:Nballs,1)
        iColorCounter<-iColorCounter+1
        #colorname<-paste("NewColor",iColorCounter)
        UrnThings$Balls[IndexOfBall]<-iColorCounter	
      }else{
        Nballs<-Nballs+1
        UrnThings$Balls[Nballs]<-WhichToCopy
      }
      #cat(WhichToCopy," ")
    }
    #return(U)
  }
  
  UrnThings$CountNumberOfColorOfFirstBall<-function(){
    ColorNeeded<-UrnThings$Balls[1]
    return (sum(UrnThings$Balls==ColorNeeded))
  }
  
  UrnThings$CountNumberOfColors<-function(){
    # counts the number of (non-black colors in the urn)
    NumColors<-length(table(Urn$Balls))
    return (NumColors)
  }
  
  UrnThings$FindCommonestColor<-function(){
    Commonest<-max(table(Urn$Balls))
    return (Commonest)
  }
  
  ###############################################################################
  ################### End of class functions ####################################
  ###############################################################################

  
  
  
  UrnThings<-list2env(UrnThings)     # this is needed, but I don't know why!	
  class(UrnThings)<-"Urn"   # the name of the class we have created

  return(UrnThings)
}
```

Having defined a new Urn class, we can define the way some common functions will work on it.
Here, we define how it will be printed.
```{r print}
print.Urn <- function(U){
  if (class(U)!="Urn") stop();
  cat("Balls: ")
  cat(U$Balls)
  cat("\nStarting configuration: ")
  cat(U$StartingConfiguration)
  cat(paste("\nUse Black ball?: ",U$UseMutationBall,"\nBlack ball weight: ",U$MutationBallWeight))
}
```


Let's test it out. First, we set up the scenario we are considering.
We will run the model until there are 50 non-black balls in the urn.
We start by considering the case in which the black ball has weight 1.
```{r scenario}
NoReps <- 1000   # how many samples to generate
WeightOfBlackBall <- 1
NoOfBalls <- 50
ThisUrn<-Urn(c(1,1),NoOfBalls,1,1)   # The urn we will use for the simulations
ThisUrn$MutationBallWeight<-WeightOfBlackBall  # now the black ball has the required weight
```


Now simulate a large number of Urns and look at the distribution of how 
many colors there are at the end.
```{r sims}
NumberOfColorsAtEnd <- rep(0,NoReps)
#pb <- progress_bar$new(total = NoReps)   # This sets=up the progress bar
for (k in 1:NoReps){
 # pb$tick()   # this makes the progress bar 'tick over'
  ThisUrn$RunUrnModel()   # simulate the urn using this weight
  NumberOfColorsAtEnd[k] <- length(table(ThisUrn$Balls))
}
```

We will now have NoReps accepted weights, so plot a histogram to see what they look like
```{r plot, echo=FALSE}
cat("\nFor black ball weight ",WeightOfBlackBall," Mean number of colors is ",mean(NumberOfColorsAtEnd))
cat("\nTable showing distribution of colors at end")
table(NumberOfColorsAtEnd)
barplot(table(NumberOfColorsAtEnd),main=paste("Distribution of number of colors at end; Black ball wt = ",WeightOfBlackBall))
```


Repeat for other weights. This shows an example of how to re-use an existing chunk of code (here the chunk that plotted the results). Rmd will insert the code in the chunk called "plot" where you put the reference "<<plot>>". (This only seems to work when you "knit" the entire file; if, instead, you "Run all" it returns an error message, for reasons I do not understand.)
```{r, echo=FALSE}
WeightOfBlackBall <- 2
ThisUrn<-Urn(c(1,1),NoOfBalls,1,1)   # The urn we will use for the simulations
ThisUrn$MutationBallWeight<-WeightOfBlackBall  # now the black ball has the required weight
for (k in 1:NoReps){
 # pb$tick()   # this makes the progress bar 'tick over'
  ThisUrn$RunUrnModel()   # simulate the urn using this weight
  NumberOfColorsAtEnd[k] <- length(table(ThisUrn$Balls))
}

 <<plot>>

WeightOfBlackBall <- 3
ThisUrn<-Urn(c(1,1),NoOfBalls,1,1)   # The urn we will use for the simulations
ThisUrn$MutationBallWeight<-WeightOfBlackBall  # now the black ball has the required weight
for (k in 1:NoReps){
 # pb$tick()   # this makes the progress bar 'tick over'
  ThisUrn$RunUrnModel()   # simulate the urn using this weight
  NumberOfColorsAtEnd[k] <- length(table(ThisUrn$Balls))
}

 <<plot>>
  
  WeightOfBlackBall <- 10
ThisUrn<-Urn(c(1,1),NoOfBalls,1,1)   # The urn we will use for the simulations
ThisUrn$MutationBallWeight<-WeightOfBlackBall  # now the black ball has the required weight
for (k in 1:NoReps){
 # pb$tick()   # this makes the progress bar 'tick over'
  ThisUrn$RunUrnModel()   # simulate the urn using this weight
  NumberOfColorsAtEnd[k] <- length(table(ThisUrn$Balls))
}

 <<plot>>
```

